---
title: "Exercise 2"
author: "Alexander Linus Grübling"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r loadData, results='hide'}
load("building.RData")
head(df)
str(df)
```

```{r splitData}
set.seed(12122371)
train_indices <- sample(1:nrow(df), size = floor(nrow(df) / 4))

train_data <- df[train_indices, ]
test_data <- df[-train_indices, ]

cat("Training data rows:", nrow(train_data), "\n")
cat("Test data rows:", nrow(test_data), "\n")

```

## 1) Computing the full model

```{r computeFullModel}
full_model <- lm(y~., data = train_data)
```

### 1a) Plotting Diagnostics

```{r}
fitted_values <- predict(full_model, train_data)
plot(train_data$y, fitted_values, main = "Fitted vs. Response",
     xlab = "Actual", ylab = "Fitted")
abline(0, 1, col = "red")

train_rmse <- sqrt(mean((train_data$y - fitted_values)^2))
print(train_rmse)
```
As we can see, the model already does a pretty good job in predicting the response. This is also visible by the comparatively low reported RMSE.

### 1b) Cross Validation

```{r cvModelEvaluation, warning=FALSE}
if (!require("cvTools", character.only = TRUE)) {
    install.packages("cvTools")
}
library(cvTools)
cv_results <- cvFit(full_model, data = train_data, y = train_data$y, 
                    cost = rmspe, K = 5, R = 100)
plot(cv_results, main = "5-fold CV with RMSE")

```
We can see, that the errors are on average pretty low. However, it is clearly visible that a number of outliers heavily influences the prediction performance.

### 1c) Cross Validation with different cost function

```{r, warning=FALSE}
cv_results_rtmspe <- cvFit(full_model, data = train_data, y = train_data$y, 
                           cost = rtmspe, K = 5, R = 100)
plot(cv_results_rtmspe, main = "5-fold CV with rtmspe")

```
By using the RTMSE instead of the RMSE, we are trimming out the outliers, and therefore get a much more even distribution of the errors.

### 1d) Predicting the Test Data

```{r}
test_predictions <- predict(full_model, test_data)
plot(test_data$y, test_predictions, main = "Predicted vs. Response",
     xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")

test_rmse <- sqrt(mean((test_data$y - test_predictions)^2))
print(test_rmse)

```
We can clearly see in this plot, that our model is predicting some values very far from their ideal values.

## 2) Best Subset Regression

### 2a) A simple way of reducing predictors
```{r}
library(dplyr)
correlation_matrix <- sapply(df, function(x) cor(x, df$y, use = "complete.obs"))
correlation_df <- data.frame(variable = names(correlation_matrix), 
                             correlation = as.numeric(correlation_matrix))

top_predictors <- correlation_df %>%
  dplyr::arrange(desc(abs(correlation))) %>%
  dplyr::slice(2:51) %>%  
  dplyr::pull(variable)

df_shortened <- df[, top_predictors]
df_shortened <- cbind(y = df$y, df_shortened)

df_full <- data.frame(response_variable = df_shortened$y, df_shortened)

models <- reg_fit <- leaps::regsubsets(y ~ ., data = df_full, nbest = 1, really.big = TRUE)
summary(reg_fit)
```
To cut down our predictor variables to just 50, I have computed the correlations each of the variables have with the response variable. Then I just used the 50 highest scoring variables to compute the best subset regression.

### 2b) Plotting the resulting object
```{r}
plot(models)
```
This plot shows the variables that best performed in the best subset regression.

### 2c) Computing the final linear model
```{r}
final_model <- lm(
  y ~ PhysFin8 + COMPLETION.YEAR + Econ14.lag3 + Econ14.lag1 + Econ14 + Econ15.lag1 + Econ3.lag3, 
  data = train_data
  )
cv_results_final_model <- cvFit(final_model, data = train_data, y = train_data$y, 
                           cost = rtmspe, K = 5, R = 100)
plot(cv_results)
```
Even though we used the RTMSE, our model still produces some heavy outliers.

### Predicting the response values
```{r}
test_predictions_final <- predict(final_model, test_data)
plot(test_data$y, test_predictions, main = "Predicted vs. Response",
     xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")

test_rmse_final <- sqrt(mean((test_data$y - test_predictions_final)^2))
print(test_rmse_final)
```
As we can see here, we have obtained a much higher RMSE score, indicating that the model is performing worse than the original one.
