---
title: "Exercise 11"
author: "Alexander Linus Grübling"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cars}
library(e1071)
library(caret)
library(ggplot2)
library(dplyr)
set.seed(12122371)
```

```{r pressure, echo=FALSE}
data <- read.csv("bank.csv", header = TRUE, sep = ";")
data$y <- as.factor(data$y)

trainIndex <- createDataPartition(data$y, p = 2/3, list = FALSE, times = 1)
trainData <- data[trainIndex, ]
testData <- data[-trainIndex, ]
```

## a)

```{r}
svm_model <- svm(y ~ ., data = trainData)
svm_predictions <- predict(svm_model, testData)

confusion_matrix <- table(Predicted = svm_predictions, Actual = testData$y)
print(confusion_matrix)

balanced_accuracy <- function(conf_matrix) {
  sensitivity <- conf_matrix[2, 2] / (conf_matrix[2, 2] + conf_matrix[1, 2])
  specificity <- conf_matrix[1, 1] / (conf_matrix[1, 1] + conf_matrix[2, 1])
  return((sensitivity + specificity) / 2)
}

bal_acc <- balanced_accuracy(confusion_matrix)
print(paste("Balanced Accuracy:", bal_acc))
```

We can see that we achieve a high count of correct "no"-class predictions. However, correct prediction of the "yes"-class are more sparse, which is expectable since it is drastically underrepresented.

## b)

```{r}
param_grid <- tune.svm(y ~ ., data = trainData,
                       gamma = c(0.001, 0.01, 0.1, 1),
                       cost = c(0.1, 1, 10, 100))

summary(param_grid)
plot(param_grid)
```

We can see that the best parameters that were found in the search grid are $\gamma = 0.1$ and $\text{cost}=1$. This trend is also visible in the plot.

## c)

```{r}
best_model <- param_grid$best.model
pred_best <- predict(best_model, testData)

conf_matrix_best <- table(Predicted = pred_best, Actual = testData$y)
conf_matrix_best

balanced_accuracy_best <- mean(
  sensitivity(conf_matrix_best),
  specificity(conf_matrix_best)
)
balanced_accuracy_best
```

With this best model found above, we achieve a dramatically better balanced accuracy.  The weakness still lies in "yes"-instances being predicted as "no", however we have fewer such cases.

## d)

```{r}
tune_result_weights <- tune(
  svm, y ~ ., data = trainData,
  kernel = "radial",
  class.weights = list(no = 1, yes = 5),
  ranges = list(cost = c(0.1, 1, 10), gamma = c(0.01, 0.1, 1))
)

print(tune_result_weights$best.parameters)

best_svm_model_weights <- svm(
  y ~ ., data = trainData, gamma = tune_result_weights$best.parameters$gamma, 
  cost = tune_result_weights$best.parameters$cost, 
  class.weights = tune_result_weights$best.parameters$class.weights
)

best_svm_predictions_weights <- predict(best_svm_model_weights, testData)

best_confusion_matrix_weights <- table(
  Predicted = best_svm_predictions_weights, 
  Actual = testData$y
)
print(best_confusion_matrix_weights)

best_bal_acc_weights <- balanced_accuracy(best_confusion_matrix_weights)
print(
  paste("Balanced Accuracy with Best Model and Class Weights:", 
        best_bal_acc_weights)
)
```

Sadly, the tunecontrol function was not working for me. Doing it the way I have done it above sadly results in a way worse balanced accuracy.